---
title: "1.1 Setup"
author: Dan Turner (dturner@u.northwestern.edu)
updated: 2-9-19
---

**Welcome to my tutorial on Shiny, a popular package for creating web applications that run on R code!**

Because of the structure of Shiny, I found it better to show off its capabilities using R Notebooks like this one. You will see that text (like this) can be set inline with code chunks...
```{r}
# ...like this.
```

Code chunks are run by pressing the *Run* button while the curser is in that code chunk, or by the key combination *Cmd+Shift+Enter*. Each code chunk behaves kind of like an isolated `.R` script file, but the results appear beneath the code chunk, instead of in the console.

*Try it out with the code chunk below.*

```{r RUN THIS}
songs <- read_csv("data/fight-songs.csv")  # Load a dataset
plot( sort( songs$sec_duration ) )         # Plot in ascending order
```

The plot above shows the length of 65 college fight songs in the US. If there was an error, you should make sure your working directory is set to the `shiny/` directory.

*Lastly, we will make sure your machine as all of the packages we will be using. Run the following chunk:*
```{r INSTALL PACKAGES}

# This script compares the required packages with what's installed.
# If any required packages are not installed, it installs them and their dependencies.
list.of.packages = c("shiny", "tidyverse", "readr", "plotly", "shinythemes", "lme4", "rsconnect")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages, dependencies = TRUE)

```

Once all of the packages are installed (if applicable) we are ready to begin!

## NEXT: BRIEF STARTER ON TIDYVERSE

*This tutorial is liscensed under https://creativecommons.org/licenses/by-sa/3.0/*
