---
title: "2.3 LAUNCHING A SHINY APP"
author: Dan Turner (dturner@u.northwestern.edu)
updated: 2-7-19
---

This script is all about how to put your shiny app online.


## Setup block
```{r setup}
# Packages
library(shiny)         # For building interactive apps
library(rsconnect)     # For connecting RStudio to shinyapps.io
library(shinyBS)       # Optional: bootstrap the shiny app for mobile 
library(shinyLP)       # Optional: easy landing pages for your shiny apps

# Other setup
null_server <- function(input, output) { } # Empty server to test different UIs with

```



##############################
## LAUNCHING YOUR SHINY APP ##
#############################

Because `shiny` is a web application that displays itself using rendered HTML, it has to be hosted on a server. Servers are just a type of specialized computer, and until now *your computer has been standing in for a server*. That means the apps you made are only visible on your machine. For other people to see it, it has to be hosted on a server connected to the internet. For this, you have two options:
     
     A. Host the app on server space that you purchase and manage (Advanced)
     
     B. Host the app on [shinyapps.io](http://www.shinyapps.io), which is managed by shiny's creators (RStudio Inc.)

As you might have guessed, it is generally more expensive to pay for shinyapps.io to manage the server than it is to manage your own, but it is fast and easy in comparison. Essentially, once you have executed the code below, it unlocks the ability to publish or deploy the app.

```{r}

# rsconnect needs three bits of information:
rsconnect::setAccountInfo(name='bias',              # the name of the account on shinyapps.io
			  token='6C22685696EC3208EC5E74CB3FA779E7',   # the token for your app
			  secret='<SECRET>')                          # the secret for your token (like a password)

# Push the current app to the shinyapps.io server, using the stored defaults
deployApp(launch.browser = TRUE, # After deploying the app online, show it in the browser
          appName = "My first shiny app") # I like to name my app this way. Otherwise, shiny uses the directory name as a default.
```
			  
When you have rsconnect configured, you just need to use deployApp. It will take your shiny app (as "app.R", for example) and upload it to shinyapps.io to a server that's been preconfigured to run it. There are a lot of formulations that `deployApp()` can take, including splitting up your code as "ui.R" for the UI and "server.r" for the server. All formulations work the same, so it's up to you whether you want to keep your UI and server in the same R script or not.



########################
## ADD A LANDING PAGE ##
########################

I will end with a last tip on deploying your shiny app. Often, you will want to have information about the application, like a guide on how to use it, or where to find more information about the work being displayed. I recommend using the shinyLP package for this. you can see a working example with the one-liner below.

```{r message=FALSE, warning=FALSE}
shinyLP::runExample()
```

# Thank you!
